---
title: "Visualize office hours"
author: "Robin Donatello"
date: "August 28, 2018"
output: html_document
---
Connect to google sheets and download the responses from the survey form. Rename the columns and shorten the class names to just MATH XXX. 
```{r, warning=FALSE, message=FALSE, results='hide'}
knitr::opts_chunk$set(echo = TRUE, warning=FALSE, message=FALSE)
library(googlesheets4)
library(dplyr)
oh_data <- "https://docs.google.com/spreadsheets/d/17XeLwLmpbfrGgKMaRriyLHIyXn6J0XT9v8tmz45CZek/edit#gid=1807123467"
# read the data into a data frame
raw <- read_sheet(oh_data)
raw[] <- lapply(raw, function(x) { attributes(x) <- NULL; x })
names(raw) <- c("ts", "8am", "9am", "10am", "11am", "12pm", "1pm", "2pm", "3pm", "4pm", "5pm", "6pm", "7pm", "8pm")
ds_raw <- select(raw, -ts)
# truncate class names
#ds_raw$class <- substr(ds_raw$class, 1, 8)
```

Reshape the data to long format, so there is not one column per time selection. Specify the order of factors so that they print out in a sensible order. 
```{r}
library(tidyr)
dsl <- gather(data=ds_raw, key=time, value=days) %>% separate_rows(days, sep=", ")
#dsl <- gather(data=ds_raw, key=time, value=days) %>% separate_rows(days, sep=", ")
# https://stackoverflow.com/questions/13773770/split-comma-separated-column-into-separate-rows/31514711#31514711 
dsl$days <- factor(dsl$days, levels=c("Monday", "Tuesday", "Wednesday", "Thursday", "Friday", "Saturday", "Sunday"))
dsl$time <- factor(dsl$time, levels=c("8pm", "7pm", "6pm", "5pm", "4pm", "3pm", "2pm", "1pm", "12pm", "11am", "10am", "9am", "8am"))
```
Count the number of records/responses per day/time. Also further stratify by class. 
```{r}
library(ggplot2); library(RColorBrewer)
all <- dsl %>% group_by(days, time) %>% summarise(n=n()) %>% na.omit()
ag <- dsl %>% group_by(days, time) %>% summarise(n=n()) %>% na.omit()
```

Create an interactive heatmap plot that shows which time slots have the highest number of students available. 
```{r, fig.width=10}
library(plotly)
 a <- ggplot(data = all, aes(x = days, y = time)) + theme_minimal()+ 
           geom_tile(aes(fill = n)) +
           scale_fill_distiller(palette="Blues", trans="reverse")
 b <- ggplot(data = ag, aes(x = days, y = time)) + theme_minimal()+ 
           geom_tile(aes(fill = n)) + 
           scale_fill_distiller(palette="Greens", trans="reverse")
ggplotly(a)
ggplotly(b)
```
